# Reading {#reading}
In this section, we learn how to read data from different file format.

## CSV

Reading csv with column headers and separated by `,`. These parameters are also the default values for the `read.csv` function.

```{r reading-csv-1, eval=F}
data <- read.csv(file = '/path/to/csv', header = TRUE, sep = ',')
```

```{r reading-csv-2, tidy=F, out.width='80%'}
# Example
data <- read.csv(file = 'data/eds.data.hurricane.csv', header = TRUE)
```

## Excel

The main advantage of Excel files is that they can store multiple tables. But reading these tables at once is different from a CSV. For this example, we're going to use the `readxl` package from the `tidyverse` collection. Please visit this [website] (https://www.tidyverse.org/) to learn more about `tidyverse`.

To read an excel file, you can use the `read_excel` function and specify at least the `path/to/the/file` and `sheet` you want to open. If you don't specify the `sheet`, `read_excel` will automatically open the first table in the spreadsheet.
In the 'eds.excel.sample.xlsx' file, there are 2 tables: heatwave and hurricane. Here's how we load both tables into R:

```{r reading-excel, eval=F}
library(readxl)
heatwave <-  read_excel(path='data/eds.excel.sample.xlsx', sheet = 'heatwave')
hurricane <-  read_excel(path='data/eds.excel.sample.xlsx', sheet = 'hurricane')
```

Once the tables are stored into individual R variable, you can perform exploration and analysis with them.

## Google Spreadsheets

If the data is stored in a Google spreadsheet, we can read it using the `googledrive` and `googlesheet4` packages. We use the `googledrive` package to log into our Google Drive account and `googlesheets4` to read the speadsheets in our drive.

In the example below, I used a spreadsheet named `eds.sample.googlesheets` which contains the same tables in the previous Excel example (heatwave and hurricane). You can clone the spreadsheet via this [link] (https://drive.google.com/open?id=1uIsgrcsevbm9voZU-rzqhTg2LE5SgEPlGabSXKTcQtc if you'd like to repeat the steps below using your Google account.

Then authenticate to your drive using `drive_auth()`. When prompted: log in, authorized googledrive, and use the authorization code if  provided. You only need to run `drive_auth()` once.

```{r reading-google-1, eval=F}
library(googledrive)
# To authenticate and authorize googledrive package
drive_auth()
```

The following scripts show how to explore a Google Drive folder. Though this is not recommended as you might encounter performance issues. 

```{r reading-google-2, eval=F}
# NOT recommended
# Then, to view the list of files in a folder
drive_ls("EDS") # where "EDS" is the folder name
# To also get the files within the subfolders
drive_ls("EDS", recursive = TRUE)
# To view the list of spreadsheets within a folder
drive_ls("EDS", type="spreadsheet")
```

Also, because of Google authentification system, you may run into an error like below when running the previous code (using `drive_ls()`). Which is why it's not recommended.

```{r reading-google-3}
#> Error in add_id_path(nodes, root_id = root_id, leaf = leaf) : !anyDuplicated(nodes$id) is not TRUE
```

To avoid this, you can use the folder url instead of the folder name. The folder url can be obtained by right-clicking on the folder and click `Get shareable link`. Then run the following code:

```{r reading-google-4, eval=FALSE}
# If using folder name doesn't work
folder_url = 'https://drive.google.com/open?id=1e0uJ9dwFcL34JA61F0tGSoaiMZ_xio_4'
drive_ls(folder_url, type="spreadsheet")
```

Then you can load the spreadsheet by using its `id`

```{r reading-google-5, eval=FALSE}
eds.sample.spreadsheet <- drive_get(id = '1uIsgrcsevbm9voZU-rzqhTg2LE5SgEPlGabSXKTcQtc')
```

It also possible to read the spreadsheet right way by using its link / `path` (without using `drive_ls()`). I recommend using this to read any Google Drive files.

```{r reading-google-6, eval=FALSE}
eds.sample.spreadsheet <- drive_get(path = 'https://drive.google.com/open?id=1uIsgrcsevbm9voZU-rzqhTg2LE5SgEPlGabSXKTcQtc')
```

Once the spreadsheet is loaded, we run a similar code used for the Excel files to read tables within the spreadsheet. But for Google Sheets, function is called `read_sheet`

```{r reading-google-7, eval=F}
library(googlesheets4)
# Authorizing the googlesheets4 package
sheets_auth(token=drive_token())
# Reading the tables
heatwave <- read_sheet(eds.sample.spreadsheet, sheet = 'heatwave')
hurricane <- read_sheet(eds.sample.spreadsheet, sheet = 'hurricane')
```

## SPSS

```{r reading-spss-1, message=F, tidy=T}
library(haven)
data <- read_sav('data/eds.spss.sample.sav')
```

By default, the `read_sav()` will read the factor levels of non-numeric and non-character variables. If, instead, we want the labels, we can run the following code:

```{r reading-spss-2, message=F, tidy=T}
library(magrittr)
library(dplyr)
# Applying haven::as_factor() to labelled columns
# Here, we already know that variables Zone, Q4 and Q50 are not factor variables.
data %>% mutate_at(vars(-Zone, -Q4, -Q50), as_factor)
```

Because variables can be labelled in SPSS, we can use them as well to find out what each column represents.

```{r reading-spss-3, tidy=T}
# To get the labels of the variables / columns
as.vector(unlist(lapply(data, function(x) attributes(x)$label)))
```

To learn more about the `haven` package and how the variables are stored, please visit: https://haven.tidyverse.org/

