# Cleaning {#cleaning}

Data cleaning is the process of getting the data ready for statistical analysis. In contrast to "Structuring the data" the target anomalies in this case are the variable values such as missing values, outliers, distribution, etc.

## Fixing skewed distribution

A data is skewed when it's distribution is not symetrical but rather distored on the left or right. Sometimes, to facilitate statistical analysis, we need to transform that skewed data so that it becomes normally distributed instead.

```{r skewed-1, echo=F, message=F}
library(haven)
library(dplyr)
data <- read.csv('data/hurricane.csv', header=T)
data <- filter(data, !is.na(data$nstorm))
hurricane <- data
```

```{r skewed-2}
# Example of skewed data
{x <- data$nstorm
h <- hist(x, breaks=20, col="grey")
xfit <- seq(min(x), max(x), length=40)
yfit <- dnorm(xfit, mean=mean(x), sd=sd(x))
yfit <- yfit*diff(h$mids[1:2])*length(x)
lines(xfit, yfit, col="red", lwd=2)}

# Log-transformation
{log.x <- log(data$nstorm)
h <- hist(log.x, breaks = 10, col = "grey")}
```

While the previous log-transformation seem to have worked, it is not entirely correct because the variable `nstorm` has `0` values. In this case, one of the commonly used method is **square root**.

```{r skewed-3}
# sqrt transformation
{sqrt.x <- sqrt(data$nstorm)
h <- hist(sqrt.x, breaks = 10, col = "grey")}
```

## Treating outliers

Outliers are usually extreme values in the datasets. If not detected and handled appropriately, they can affect the accuracy of the predicitions and analysis. Treating outliers depends on a good knowledge of the data and it is up to the data analyst to decide on how to go about on fixing them. [Selva Prabhakaran](http://r-statistics.co/Outlier-Treatment-With-R.html) gave some ways of handling outliers that we're going to show below.

You can detect outliers using `boxplot` and they will show up as dots outside the whiskers:

```{r outliers-1}
# For continuous variable
outlier_values <- unique(boxplot.stats(data$nstorm)$out)  # outlier values.
boxplot(data$nstorm, main="How many storm have you experienced?", boxwex=0.1)
mtext(paste("Outliers: ", paste(outlier_values, collapse=", ")), cex=0.6)
```

```{r outliers-2}
# For categorical variable
boxplot(nstorm ~ worry, data=data, main="Number of storms experienced across levels of worry")
```

Approaches for outlier treatment include imputation with `mean`, `median`, or `mode`. We can also cap the values, predict them, or delete the observations. We'll talk more about that below.

## Fixing missing values

Ways of detectng missing values (`NA`) include the `summary` and `anyNA()`.

```{r missing-1, include=F}
x <- sample(1:nrow(data), 50, F)
data[x,'nstorm'] <- NA
data <- as_tibble(data)
```

```{r missing-2}
anyNA(data$nstorm)
summary(data$nstorm)
```

The easiest way to handle them is by removing all of the corresponding observations using `tidyr::drop_na()`. Or, in some cases, removing the variable itself.

```{r missing-3}
tidyr::drop_na(data)
```

However, if dropping all of the rows with missing values affect the quality of the data, then another option is to replace the missing values with the mean/median/mode of the variable or predict using an appropriate algorithm. There are several packages out there that are solely dedicated to treating missing values including `VIM` and `MICE`.

In this next example, we'll try to predict the 15 missing values in the variable `nstorm` (number of storms the survey respondents have experienced) using the variables that has no missing values: `zone`, `lat`, and `long`.

```{r missing-4, message=F}
# Imputation using MICE
library(mice)

# Building the mice model
mice_model <- mice(select(data, zone, lat, long, nstorm), method="rf", printFlag=F)
# Predicting the missing values
# generate the completed data
mice_prediction <- complete(mice_model)
# checking for NAs
anyNA(mice_prediction)
```

Then we can visualize the data to see how well the imputation has performed. However, the best way to assess the accuracy is to compare actual values with predicted values using measures such as: `MSE`, `MAE`, `MAPE`, etc.

```{r missing-5, message=F}
# Visualizing the prediction
non_na_latitude <- data$lat[!is.na(data$nstorm)]
non_na_nstorm <- data$nstorm[!is.na(data$nstorm)]
na_latitude <- mice_prediction$lat[is.na(data$nstorm)]
na_nstorm <- mice_prediction$nstorm[is.na(data$nstorm)]
plot(non_na_nstorm, non_na_latitude, col="grey", pch="•", ylab="Latitude", xlab="Number of Storms Experienced")
points(na_nstorm, na_latitude, col="red", pch="•", cex=2)
legend("topright", c("Existing values", "Predicted missing values"), col=c("grey", "red"), pch="•", cex=1.5)
```

Other ways of imputing the missing values are with `mean`, `median`, or mode.

```{r missing-6}
prediction <- data
na.observations <- is.na(data$nstorm)
prediction$nstorm[na.observations] <- median(prediction$nstorm[!na.observations])
# Visualizing the prediction
na_latitude <- prediction$lat[na.observations]
na_nstorm <- prediction$nstorm[na.observations]
plot(non_na_nstorm, non_na_latitude, col="grey", pch="•", ylab="Latitude", xlab="Number of Storms Experienced")
points(na_nstorm, na_latitude, col="red", pch="•", cex=2)
legend("topright", c("Existing values", "Predicted missing values"), col=c("grey", "red"), pch="•", cex=1.5)
```

For mode, there is no built-in function in R but I found one [here](https://www.tutorialspoint.com/r/r_mean_median_mode.htm):
```{r missing-7}
getmode <- function(v) {
   uniqv <- unique(v)
   uniqv[which.max(tabulate(match(v, uniqv)))]
}
```

