---
output:
  pdf_document: default
  html_document: default
---
# Analysis {#analysis}

## Working with survey data

We'll use a survey [data](https://osf.io/79t2k/) from a study of approach to engage christians in the issue of global warming. Using this data allow us to compare our results with the original. In this exercise, we'll replicate the application of sampling weights and the standardization of variables which are common treatments of raw survey data.

```{r survey-data-1, echo=F, warning=F, message=F}
library(dplyr)
library(haven)
library(stringr)
library(rvest)
library(tidyr)
library(survey)
rm(list=ls())

data <- read_sav('https://files.osf.io/v1/resources/79t2k/providers/osfstorage/5c6c390e82a3950018c77ec2?action=download&direct&version=1')
# getting raw data
data <- data %>% select(-starts_with('Z')) %>% filter(`filter_$`==1)
# getting subset of variables
data.pre <- data %>% select(ends_with('_pre'))
data.post <- data %>% select(ends_with('_post'))
data.others <- data %>% select(-ends_with('_pre'), -ends_with('_post'))
selected.cols <- c(names(data.others)[c(1,2,4,10,11,21,13)], names(data.pre)[1:3], names(data.post)[1:3])
data <- data %>% select(one_of(selected.cols))
cat('Selected variables')
data.variables <- cbind(
  names(data), sapply(data, function(x)attr(x, 'label'))
  ) %>% as_tibble()
data.variables$V2 <- format(data.variables$V2 , justify = "left")
data.variables %>% print()
```

### Variable Standardization

Standardizing variables must be performed before any regression or other statistical analysis are done with the data. It can be done by rescaling the variables using the z-score formula. Standardized variables will then have a mean of 0 and a standard deviation of 1.

```{r survey-data-2, warning=F, message=F}
# z-score formula
get.z.score <- function(X){
  (X-mean(X, na.rm=T))/sd(X, na.rm=T)
}
# standardizing the pretest and posttest variables
Zdata <- data %>% mutate(across(ends_with('_pre')|ends_with('_post'), get.z.score, .names='Z{col}'))
Zdata
```

### Sample weighting

Per Josep Espaga Reig's [book](https://bookdown.org/jespasareig/Book_How_to_weight_a_survey/introduction.html), there are 4 steps in survey weighting:

- Base/design weights: if participants have different probability of being sampled;
- Non-response weights: if some participants didn't respond to the survey;
- Use of auxillary data/calibration: adjusting weights to the total population;
- Analysis of weight variability/trimming: to check variability in the computed weights;

In this example, we will estimate of the percent of US christian population who either believe global warming (GW) is happening or worried about GW. We will give descriptives of the distribution of these two variables and then a simple extrapolation and compute total for the whole US christian population.

Our variables of interest are:

- **GWHap**: On a scale from 1 to 7, how strongly do you believe that global warming is or is not happening?		
- **GWworry**: How worried are you about global warming?

#### Step 1: Design weights

In the [paper](https://www.researchgate.net/publication/334298965), it is mentionned that participants were recruited via Prime Panels. And only those who previously identified as christians were selected. It should be good to assume that each participants had an equal probability of being sampled. Therefore, _we can skip the Step 1 of the weighting_. If we had to compute the design weight $d_i$, it would be equal to the inverse of the probability of being sampled: $1/p_i$. Per Josep Espaga Reig's [book](https://bookdown.org/jespasareig/Book_How_to_weight_a_survey/introduction.html): the design weights can be interepreted as "the number of units in our population that each unit in our sample represents. [...] The sum of all design weights should be equal to the total number of units in our population".

#### Step 2: Non-response weights

The paper also doesn't mention any information about non-respondent during the survey. So we can assume that there is no bias in the responses and _we can skip Step 2_. In short, the goal of Step 2 is to account for differences in propensity to respond among the participants. For example, participants from specific neighborhood, income level, etc may have lower response rates, and all the response we have are from a specific demographic groups, which would bias the analysis.

#### Step 3: Using auxillary data for weight calibration

The goal here is to calibrate the survey data to the population in general. So we would need information on both the survey respondents and the population. In this case, we will use demographic variables (gender, education, and race). One of the most common method for calibration weighting is called: _raking_ which we will use as an the example.

We'll use as auxiliary data the [PEW Datasets](https://www.pewforum.org/data/) which gives some statistics of the US christian population. We'll scrape the data directly from their website so that we'll have fewer data wrangling to do.

```{r survey-data-3, warning=F, message=F}
# PEW data
# getting only the christian respondent and their weights
pew.data <- read_sav('data/pew-data.sav') %>% 
  select(RELTRAD, agerec, SEX, racethn, WEIGHT) %>% 
  mutate(across(where(is.labelled), as_factor)) %>% 
  mutate(across(where(is.factor), as.character)) %>% 
  mutate(
    religion = RELTRAD %>% recode(
      `Evangelical Protestant Tradition`='christian',
      `Mainline Protestant Tradition`='christian',
      `Historically Black Protestant Tradition`='christian',
      `Catholic`='christian',
      `Mormon`='christian',
      `Orthodox Christian`='christian',
      `Jehovah's Witness`='christian',
      `Other Christian`='christian',
      `Jewish`='other',
      `Muslim`='other',
      `Buddhist`='other',
      `Hindu`='other',
      `Other World Religions`='other',
      `Other Faiths`='other',
      `Unaffiliated (religious "nones")`='other',
      `Don't know/refused - no information on religious identity`='other',
    ),
    pew.age = agerec %>% recode(
      `Age 24 or younger`='18-29',
      `Age 25-29`='18-29',
      `30-34`='30-49',
      `35-39`='30-49',
      `40-44`='30-49',
      `45-49`='30-49',
      `50-54`='50-64',
      `55-59`='50-64',
      `60-64`='50-64',
      `65-69`='65+',
      `70-74`='65+',
      `75-79`='65+',
      `80-84`='65+',
      `85-89`='65+',
      `Age 90 or older`='65+',
      `Don't know/refused`='NA'
    ),
    pew.race = racethn %>% recode(
      `White non-Hispanic`='White',
      `Black non-Hispanic`='Black',
      `Hispanic`='Hispanic',
      `Other`='Other',
      `Don’t know/Refused (VOL.)`='NA'
    )
  ) %>% 
  filter(religion=='christian') %>% 
  select(SEX, pew.age, pew.race, WEIGHT) %>% 
  na.omit()

names(pew.data)[1] <- 'pew.gender'

head(pew.data,10)

pew.gender_age <- pew.data %>% 
  select(pew.age, pew.gender, WEIGHT) %>% 
  group_by(pew.gender, pew.age) %>% summarise(n=sum(WEIGHT)) %>% 
  pivot_wider(names_from=pew.age, values_from=n) %>% 
  select(-`NA`)

pew.race <- pew.data %>% select(pew.race, WEIGHT) %>% 
  group_by(pew.race) %>% summarise(n=sum(WEIGHT)) %>% 
  filter(pew.race!='NA')
```

Now we'll recode our age and race variables to match the PEW data.

```{r survey-data-4, warning=F, message=F}
# interaction between gender and age
our_data.recoded <- data %>% 
  select(ex_ppage, ex_ppgender, raceCat) %>% 
  mutate(across(where(is.labelled), as_factor)) %>% 
  mutate(
    age_group=cut(ex_ppage %>% as.numeric(), breaks=c(18, 30, 50, 65, 100), right=F)
  ) %>% 
  mutate(across(where(is.factor), as.character)) %>% 
  mutate(
    age_group=age_group %>% recode(
      `[18,30)`='18-29',
      `[30,50)`='30-49',
      `[50,65)`='50-64',
      `[65,100)`='65+'
    ),
    race_cat=raceCat %>% recode(
      `White Non-hispanic`='White',
      `Black Non-hispanic`='Black',
      `Hispanic`='Hispanic',
      `Other Non-hispanic`='Other',
      `Two+ races non-hispanic`='Other'
    )
  ) %>% 
  unite(col=gender_age, ex_ppgender, age_group, remove=F) %>% 
  mutate(gender_age=replace(x=gender_age, list=gender_age %in% c('Female_NA', 'Male_NA'), values=NA)) %>% 
  na.omit()

# number of participants/weights by units
# this would be the sum of the final weights from Step 1 and 2 if they were necessary
our_data.gender_age <- our_data.recoded %>% 
  group_by(age_group, ex_ppgender) %>%
  summarise(n = n()) %>% 
  pivot_wider(names_from=age_group, values_from=n)

our_data.race <- our_data.recoded %>% 
  group_by(race_cat) %>%
  summarise(n = n())

rm(list=setdiff(ls(), c('data', 'data.variables', 'Zdata', 'pew.data', 'pew.gender_age', 'pew.race', 'our_data.recoded', 'our_data.gender_age', 'our_data.race')))
```

Now, we'll scale the PEW data to our sample size.

```{r survey-data-5, warning=F, message=F}
# our total (weighted) observations
our_data.wgt <- nrow(our_data.recoded)

# scaling PEW to our data
# for age and gender
pew.gender_age.scaled <- pew.gender_age %>%
  pivot_longer(-pew.gender, names_to='pew.age', values_to='WEIGHT') %>%
  unite(col=gender_age, pew.gender, pew.age)

total.pew <- pew.gender_age.scaled$WEIGHT %>% sum()

pew.gender_age.scaled %<>%
  mutate(Freq = round(WEIGHT/total.pew * our_data.wgt, 0) ) %>%
  select(-WEIGHT)

pew.gender_age.scaled

# for race
pew.race.scaled <- pew.race %>%
  mutate(
    Freq = round(n/total.pew * our_data.wgt, 0),
    race_cat=pew.race
  ) %>%
  select(-n, -pew.race)

pew.race.scaled
```

**Implementing calibration**
We'll use the same procedure explained in this [book](https://bookdown.org/jespasareig/Book_How_to_weight_a_survey/calibration.html) which uses the `survey` package.

```{r survey-data-6, warning=F, message=F}
# No weights or probabilities supplied, assuming equal probability
our.svydesign <- svydesign(ids = ~ 0, data=our_data.recoded)

# the variable used for calibration are: gender/age interaction and race
our.raked <- rake(our.svydesign, sample.margins = list(~race_cat, ~gender_age), population = list(pew.race.scaled, pew.gender_age.scaled))

# collecting the weights
raked.weight <- our.raked$postStrata[[1]][[1]] %>% attributes() %>% .[["weights"]]
weighted_data <- our_data.recoded
weighted_data$raked.weight <- raked.weight

head(weighted_data, 10)
```


```{r include=F}
# scraping data from PEW
url = 'https://www.pewforum.org/religious-landscape-study/christians/christian/'
pew.page <- read_html(url)
get_pew_table <- function(page, table_i){
  header <- html_text(html_nodes(page, "table.highchart thead.right-aligned")[table_i])
  content <- html_text(html_nodes(page, "table.highchart tbody")[table_i])
  h.rm.t <- gsub('\\t+',';',header); h <- strsplit(gsub('\\n+','',h.rm.t), ';')[[1]]
  c.rm.t <- gsub('\\t+',';',content); c <- strsplit(gsub('\\n+','',c.rm.t), ';')[[1]]
  hl <- length(h); cl <- length(c); r <- cl/hl;
  c.split <- split(c, rep(1:r, rep(hl, r)))
  pew.table <- as_tibble(do.call(rbind, c.split)); names(pew.table) <- h
  pew.table <- pew.table %>% mutate_all(funs(str_replace_all(.,'%|,',''))) %>% 
    mutate_all(as.integer) %>% mutate_all(funs(.*.01*320e6*.706))
  return(pew.table)
}
```















